---
title: "Exercise #1"
subtitle: "Fortgeschrittene Statistische Software für NF - SS 2024"
date: "`r Sys.Date()`"
author: "Pat Callahan (Matrikelnummer: 12672775)"
output:
  bookdown::pdf_document2:
    extra_dependencies: ["float"]
  bookdown::html_document2:
    base_format: distill::distill_article
    toc: true
    toc_depth: 2
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r setup, include=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      options(dplyr.print_max = 5),
                      fig.pos = "H", out.extra = "")
library(dplyr)
library(lubridate)
library(magrittr)
library(ggplot2)
```

\newpage

# Exercises

## Exercise 1: Car Rentals (5 Points)

The file `"car_rentals.xlsx"` (on moodle) contains information about
rented cars. Load it in R and into your environment.

```{r}
car_rentals <- readxl::read_excel("car_rentals.xlsx")
```

### Combine the columns containing rental date information to a column named "rental_date". Use two different functions to assign the date variable class

```{r}
(car_rentals %<>%
  tidyr::unite("rental_date", 
               rental_date_year, rental_date_month, rental_date_day, 
               sep = "-") %>%
  # mutate(rental_date = as.Date(rental_date)) 
  mutate(rental_date = ymd(rental_date)) )
```

### Create a visualization that makes use of the newly created date variable and at least one other variable.

```{r cum-travel-plot, fig.cap="From sample data set"}

car_rentals %>%
  group_by(purpose) %>%
  mutate(cumulative_miles = cumsum(total_km)) %>%
  ggplot(aes(rental_date, cumulative_miles, color = purpose)) +
  geom_line() +
  scale_color_manual(labels = c("Business", "Private"), values = c("red", "blue")) +
  scale_x_date(date_breaks = "1 month", date_labels =  "%b %Y") +
  theme_bw() +
  labs(title = "Miles Travelled between Oct. '22 and Mar. '23", 
       subtitle = "Cumulative Kilometers Grouped by Travel Purpose",
       x = "Date",
       y = "Cumulative\nKilometers",
       color = "Travel Purpose") +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.title.y = element_text(angle = 0, vjust = 0.5),
    axis.text.x = element_text(angle = 25, vjust = 0.5),
    panel.grid.minor = element_blank()
    )

```

Interpretation of Figure \@ref(fig:cum-travel-plot): in the given time
interval, the amount of miles travelled for both business and private
purposes has been increasing at a mostly linear trend (with the
exceptions of the ends of the private travel), and there is consistently
more distance being travelled for private purposes.

### Create a new column that contains the difference in time between car rental and car return

```{r}
(car_rentals %<>%
  tidyr::unite("return_date", 
               return_date_year, return_date_month, return_date_day, 
               sep = "-") %>%
  mutate(return_date = ymd(return_date)) %>%
  mutate(rental_duration = difftime(return_date, rental_date)))
```

### Use tidyverse functions to calculate the average rental duration grouped by the purpose of rental. Interpret your results

```{r}
car_rentals %>%
  mutate(rental_duration = as.numeric(rental_duration)) %>%
  group_by(purpose) %>%
  summarise(
    mean_rental_duration_days = round(mean(rental_duration), 2)
  )
```

The average rental duration is nearly 3 days more for private purposes
than for business purposes, and this may explain a large portion of the
difference in cumulative miles travelled for these different purposes.

## Exercise 2: Programming Languages (4 Points)

Download the languages dataset from moodle and load it into R. The data
is provided by the programming language database
[PLDB](https://pldb.io/). You can find a detailed description of all
variables in the data on the PLDB's website at
<https://pldb.io/csv.html> and a short description below.

```{r}
languages <- readr::read_csv("languages.csv")
# glimpse(languages) 
```

### Generate a table with the top 15 programming languages and print the language's official title, the year it first appeared and the names of its creators using `knitr::kable()`.


```{r capt}
languages %>%
  slice_min(language_rank, n = 15) %>%
  select(Title = title, "Inception Year" = appeared, Creators = creators) %>%
  knitr::kable(
    caption = "Top 15 Most Popular Programming Languages (Source: PLDB)"
  )
```


### Add a caption to your table. Who created the second most famous programming language?

```{r}

# Note that language_rank indexes at 0!
languages %>%
  dplyr::filter(language_rank == 1) %>%
  pull(creators)
```

### Calculate the age of each programming language based on the year it first appeared. Automatically use the current year, so that your code will still correctly function next year. Add the column to the dataframe.

```{r}
(languages %<>%
  mutate(language_age = year(Sys.Date()) - appeared))

languages %>%
  select(title, language_age) %>%
  head(5)
```

### Create a scatterplot, showing the estimated number of users of a programming language on the X-axis and the number of papers about the language on the Y-axis.

See below.

### Add an appropriate caption to your figure (as part of the Rmarkdown, not within ggplot) and align the figure to the left of the page.

See below.

### Write a short sentence describing and interpreting the figure. Properly [reference](https://bookdown.org/yihui/rmarkdown-cookbook/cross-ref.html) the figure in your sentence.

See below.

### Write another sentence describing that the figure was generated using `ggplot2` and add a [citation](https://bookdown.org/yihui/rmarkdown-cookbook/bibliography.html) for `ggplot2`.

```{r prog-lang-plot, fig.cap="Data sourced from the Programming Language Database.", fig.align='left'}
# I don't see anything about papers, but rather a books column...
languages %>%
  ggplot(aes(x = log(number_of_users), y = semantic_scholar)) +
  geom_point(alpha = 0.4, color = "red") +
  theme_bw() +
  labs(title = "Number of Users vs. Books\nfor Various Programming Languages", 
       x = "Natural Log of Number of Users",
       y = "Semantic\nScholar\nScore") +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.title.y = element_text(angle = 0, vjust = 0.5),
    axis.text.x = element_text(angle = 25, vjust = 0.5),
    panel.grid.minor = element_blank()
    )

```

In general, Figure \@ref(fig:prog-lang-plot) indicates that the more
users a given language has, the greater the quantity of published
content is available for it; however, this association appears quite
weak and there is a great amount of variability in this. I opted to plot
the log of the number of users given that some popular languages (e.g.
SQL, JavaScript) will and do have many orders of magnitude more users
than some of the obscure, esoteric languages.

Plot generated using the {ggplot2} package [@wickham2016].

### Calculate the correlation between the estimated number of users and number of papers. Write a sentence in Markdown and automatically insert the correct number.

```{r}

(book_user_cor <- cor(languages$number_of_users, 
                      languages$semantic_scholar, 
                      use = "complete.obs"))

```

There is a correlation of `r round(book_user_cor, 3)` between the number
of publications measured by Semantic Scholar and the number of users of
a language.

## Exercise 3: Programming Languages 2 (5 Points)

### The column `number_of_users` holds the estimated number of users for each programming language. Compute the fraction of users each programming language has, based on the total sum of users in the dataset.

```{r}

total_users <- sum(languages$number_of_users)
languages %>%
  group_by(pldb_id) %>%
  summarize(
    user_fraction = number_of_users/total_users
  )

```

### Output a formatted table with the **top 13** languages ordered by the fraction of users. Include the language's title, the number of users and the fraction of total users.

-   Add a caption to the table.
-   Display the fraction of users rounded to 3 digits, without rounding
    by hand.

> Tip: Check out the options in `knitr::kable()`

```{r popular-langs}

languages %>%
  select(title,number_of_users) %>%
  slice_max(number_of_users, n=13) %>%
  group_by(title) %>%
  summarize(
    number_of_users = number_of_users,
    user_fraction = number_of_users/total_users
  ) %>%
  arrange(desc(user_fraction)) %>%
  rename(Language = title, 
         "Number of Users" = number_of_users, 
         "User Fraction" = user_fraction) %>%
  knitr::kable(digits = 3, 
               caption = "Top 13 most popular programming languages by user count",
               format.args = list(big.mark = ",")) 
```

### Write one or two sentences describing the table, focusing on its *13th* entry in particular. Include an adequate [citation](https://bookdown.org/yihui/rmarkdown-cookbook/bibliography.html) for the language and properly [reference](https://bookdown.org/yihui/rmarkdown-cookbook/cross-ref.html) the table.

> Note: Proper referencing of the table will only work when rendering to
> a PDF. Don't get hung up on this if it provides some problems, as
> dealing with LaTeX can be a bit tricky.

Table \@ref(tab:popular-langs) shows that the R programming language
[@rcoreteam2024] is the 13th most popular programming language based on
reported number of users. Given its growing popularity in university
courses, academia, and in some industries such as pharmaceuticals as a
free and open source language, it is unsurprising that this
traditionally "statistical programming language" is among the most
popular in the world.

\newpage 
# References

(It wasn't initially clear to me why the instructions requested I move
this section to before Supplementary Materials as I had only been
rendering the distill document, but I see the PDF rendering places the
references here.)

::: {#refs}
:::

\newpage

# Supplementary Materials

### Create a histogram illustrating the distribution of the ages of different programming languages.

```{r}
languages %>%
  ggplot(aes(language_age)) +
  geom_bar() +
  labs(title = "Histogram of Programming Language Ages") +
  xlab("Age of Language") +
  ylab("Number of Languages")
```

### Do you encounter anything surprising in your histogram? Briefly describe the issue at hand and explore the data to better understand why it occurs.

Surprisingly, there are programming languages with origins much older
than the 20th century. Filtering the data set to only view languages
greater than 100 years old reveals that numeric systems one would not
traditionally think of as a "computer" language in the modern sense are
included such as Morse code and Arabic numerals.

```{r}
languages %>%
  dplyr::filter(language_age > 100)
```

### Update the graphic to resolve this issue or work around it and add a caption to your figure.

I'm arbitrarily choosing to filter for languages that appeared only in
the years immediately preceding the second world war given that many
advancements towards modern computing began around this time.

```{r, fig.cap="Distribution of programming language ages, filtered for languages that emerged in 1930 or later."}

languages %>%
  dplyr::filter(appeared >= 1930) %>%
  ggplot(aes(language_age)) +
  geom_bar() +
  labs(title = "Histogram of Programming Language Ages",
       subtitle = "Created since 1930") +
  xlab("Age of Language") +
  ylab("Number of Languages")

```
